Inattentive responding can induce spurious associations between task behavior and symptom measures

### 2022-11-09 manuscripts/main.tex
checkpoint revisions progress

% specify link style
\addbibresource{main.bib}
\renewcommand*{\bibfont}{\small}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
    citecolor=blue
}

@@

% set up the author block
\setlength{\affilsep}{2em}   % set the space between author and affiliation
\author[1]{Samuel Zorowitz}
\author[1,2]{Yael Niv}
\author[1,3]{Daniel Bennett}
\affil[1]{Princeton Neuroscience Institute, Princeton University, USA}
\affil[2]{Department of Psychology, Princeton University, USA}
\affil[3]{Department of Psychiatry, Monash University, Australia}

@@

\section{Introduction}

@@

Here we wish to draw special attention to an underappreciated feature of online psychiatric research using self-report symptom surveys. In such surveys, participants rate their endorsement of various psychiatric symptoms and, since most individuals in the general population tend to endorse no or few symptoms in many symptom domains, the resulting ground-truth symptom distributions tend to be heavily positively skewed \cite{lowe2008validation, tomitaka2018distributional}. In this situation, the assumption that C/IE responding merely increases unsystematic measurement noise becomes untenable. Because of the positive skew in the ground-truth symptom distribution, participants who respond carelessly to the symptom survey are more likely to report higher levels of symptom endorsement relative to participants who complete the survey accurately \cite{chandler2020participant, ophir2020turker}. Consequently, unless C/IE survey responses are carefully identified and removed, a considerable proportion of putatively symptomatic individuals in an online sample may, in fact, be participants who have not engaged with the experiment with sufficient attention or effort.

@@

Here, we demonstrate the real risk that C/IE responding can lead to spurious symptom-task correlations in computational psychiatry research. First, we asked to what extent recent studies in computational psychiatry screen participants based on self-report symptom data. We found that the majority of these studies did not screen participants' survey data at all, and that very few followed best-practice recommendations for survey data screening. We then asked whether behavioral screening alone was sufficient to identify participants engaging in C/IE responding on psychiatric symptom surveys. In a new dataset from two separate online labour marketplaces, we found that screening based on task behavior fails to adequately identify participants engaging in C/IE responding on surveys. Lastly, we investigated whether, under these circumstances, C/IE responding led to spurious correlations between symptom severity and task performance for positively-skewed symptom measures. Consistent with the logic set out above, we confirmed that failure to appropriately screen out C/IE survey responding in the proof-of-concept dataset that we collected would have produced a number of spurious correlations between task behavior and self-reported symptoms of depression and anxiety that are abolished when data are screened properly.

@@

\begin{table}[!t]
\centering
\small
\setlength{\tabcolsep}{6pt}
\begin{tabular}{lcccccccc}
\toprule
& & \multicolumn{3}{c}{Total Score} & \multicolumn{2}{c}{Cronbach's $\alpha$} & \multicolumn{2}{c}{\% Clinical Cutoff} \\
\cmidrule(lr){3-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
 Subscale & Skew & Attentive & C/IE & \textit{t}-value & Attentive & C/IE & Before & After \\
\midrule
7-up   &  0.81 &    3.87 &   10.15 &  -13.31^* &    0.84 &    0.84 &     13.0\% &       4.0\% \\
7-down &  0.76 &    4.75 &   10.68 &   -9.99^* &    0.94 &    0.88 &     17.4\% &       9.3\% \\
GAD-7  &  0.75 &    4.86 &    9.73 &   -7.88^* &    0.92 &    0.87 &     25.9\% &      17.3\% \\
BIS    &  0.78 &    7.74 &    7.92 &   -0.54 &    0.83 &    0.62 &        - &             - \\
BAS    &  0.17 &   15.73 &   16.16 &   -0.91 &    0.84 &    0.71 &        - &             - \\
SHAPS  &  0.26 &    8.02 &   10.85 &   -4.04^* &    0.90 &    0.81 &     17.9\% &      14.6\% \\
PSWQ   &  0.19 &    4.78 &    6.74 &   -4.78^* &    0.93 &    0.81 &      7.3\% &       7.0\% \\
\bottomrule
\end{tabular}
\captionsetup{width=0.88\textwidth}
\caption{Descriptive statistics of the self-report symptom measures between attentive and C/IE participants. Skew: the empirical skewness of the distribution of total symptom scores. Total score: the average symptom score across attentive and C/IE participants. Stars indicate statistical significance at $p<0.05$. Cronbach's $\alpha$: a measure of response consistency, where values closer to 1 indicate greater consistency in responses. \% Clinical Cutoff: the percentage of participants reaching threshold for clinical symptomology before and after screening based on the infrequency measure. (The BIS/BAS scales do not have clinical thresholds.)}
\label{tab:validation}
\end{table}

@@

\subsection{C/IE responding manifests as qualitatively distinct behavioral strategy}

Finally, we employed a theory-agnostic modeling approach to investigate how C/IE participants on the probabilistic reversal-learning task compared to attentive participants. The motivation for this analysis was to better understand why C/IE responding was inconsistently predicted by chance-level performance, and also correlated with asymmetric learning rates.

To characterize participants' choice behavior, we adapted the softmax regression model from \cite{seymour2012serotonin}. This model estimates, for each participant, how much their choice depends on the recent history of trial events (rewarding outcomes, non-rewarding outcomes, and choices from the preceding 5 trials; see Methods). The regression weights for each event, averaged within attentive and C/IE participants, are presented in Figure \ref{fig:regression}. Comparing attentive to C/IE participants, we observed a credible difference (i.e. 95\% highest density intervals excluded zero) only for the $T-1$ weight for previous choice. That is, attentive participants were more likely to repeat their previous choice (i.e. greater choice hysteresis) than were C/IE participants.

\begin{figure}[!b]
\includegraphics[width=16cm]{../figures/main_03c.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Softmax regression decision weights in attentive (red) and C/IE (blue) participants. The weights dictate the extent to which the recent history of rewards, nonrewards, or previous choices influence current choice. Error bars indicate the 95\% highest density interval of each weight. * denotes where the 95\% highest density interval of the difference in weights excluded zero.}
\label{fig:regression}
\end{figure}

@@

One complicating factor for our argument is that C/IE responding may manifest in other ways than simply random responding for both behavioral tasks and self-report surveys. Indeed, there are more ways to respond carelessly than to respond attentively to a task or self-report inventory (e.g., random response selection, straight-lining, zig-zagging, acquiescence bias) \cite{curran2016methods}. Here we observed that C/IE participants failed to maintain previously correct choices after an occasional non-reward in the probabilistic reversal learning task, perhaps responding more locally based on the most immediate reward. The specific response strategy a participant adopts is likely to reflect the idiosyncratic integration of multiple perceived benefits (e.g. time saved, effort avoided) and costs (e.g. loss of performance bonuses, risk of detection and pay forfeited). As has been previously documented \cite{desimone2018dirty}, the presence of multiple response strategies makes it clear why certain screening measures are more or less likely to correlate. For example, the inter-item standard deviation and personal reliability measures are both sensitive to statistically random responding, but less sensitive to straight-lining. Most importantly, a diversity of heuristic response strategies highlights the need for many screening measures of C/IE responding, each sensitive to different heuristic strategies.

@@

One major concern with performing rigorous screening and exclusion of participants based on C/IE detection methods is that we might inadvertently introduce an overcontrol bias \cite{elwert2014endogenous}. That is, we have been treating a tendency towards C/IE responding as independent from psychopathology; however, to the extent that C/IE responding reflects lack of motivation \cite{barch2015mechanisms}, avoidance of effort \cite{cohen2001impairments, culbreth2016negative}, or more frequent lapses of attention \cite{kane2016individual, robison2017neurotic}, one might hypothesise a true underlying association between psychopathology and careless responding in online studies. In such cases, it is plausible that rigorous screening of C/IE responding might lead to the differential exclusion of truly symptomatic participants.

However, when using self-report questionnaires for screening, for overcontrol to seriously impact results it would have to be the case that symptomatic participants frequently endorse improbable or impossible responses to infrequency-item checks (e.g., responding `Agree' to ``I competed in the 1917 Olympic Games"). In this case, and even if such participants truly are experiencing severe symptoms of motivation or attention, there is likely to be limited utility in measuring these symptoms using a self-report measure that they are unable to complete veridically. A similar rationale underlies the widespread use of semi-structured interviews and other clinician-report measures rather than self-report measures for in-clinic psychiatric research. We would therefore argue that, if the psychiatric phenomenon being studied is such that this issue warrants concern, the research question may be better suited to an in-person study design involving participants in the clinic who meet full diagnostic criteria than a correlational design involving an online convenience sample.

@@

409 total participants were recruited to participate in an online behavioral experiment in late June - early July, 2020. Specifically, 208 participants were recruited from Amazon Mechanical Turk (MTurk) and 201 participants were recruited from Prolific. This study was approved by the Institutional Review Board of Princeton University (\#5291), and all participants provided written informed consent. Total study duration was approximately 10 minutes per participant. Participants received monetary compensation for their time (rate USD \$12/hr), plus an incentive-compatible bonus up to \$0.25 based on task performance.

@@

\subsubsection{Theory-agnostic choice analysis} In the final analysis, we use a theory-agnostic modeling approach to characterize the choice behavior of attentive participants and those suspected of C/IE responding. To do so, we adapted the softmax regression model from \cite{seymour2012serotonin}. The softmax regression model estimates the influence of the history of past rewards, non-rewards, and choices on a participant's current choice.

Specifically, the influence of the history of particular type of event is defined as:

where $x_{t-i}$ is a binary indicator [0,1] denoting if an event (i.e. reward, non-reward, previous choice) occurred on trial $t-i$ and $w_{t-1}$ is the associated decision weight. These weights were estimated for rewards, non-rewards, and previous choices up to five trials in the past.

The overall tendency to choose a particular choice option is dictated by a softmax choice rule:

We fit the softmax regression model using Stan (details above).Participant parameters were fit individually (i.e. not hierarchically)  so as to prevent bias during parameter estimation from partial-pooling between attentive and C/IE participants. Parameters were sampled with Gaussian priors with $\mu = 0$ and $\sigma = 5$.

@@

\printbibliography
\pagebreak

@@

\section*{Supplement}
\setcounter{figure}{0}
\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}

@@

\clearpage
\subsection*{Posterior predictive check}

>>>>

% setup references
\defbibheading{main}{}
\defbibheading{supp}{}
\addbibresource[label=main]{main.bib}
\addbibresource[label=supp]{supp.bib}
\renewcommand*{\bibfont}{\small}

% specify link style
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
    citecolor=blue
}

@@

% set up the author block
\setlength{\affilsep}{2em}   % set the space between author and affiliation
\author[1]{Samuel Zorowitz}
\author[1,2]{Yael Niv}
\author[1,3]{Daniel Bennett}
\affil[1]{Princeton Neuroscience Institute, Princeton University, USA}
\affil[2]{Department of Psychology, Princeton University, USA}
\affil[3]{School of Psychological Sciences, Monash University, Australia}

@@

\begin{refsection}[main]
\section{Introduction}

@@

Here we wish to draw special attention to an underappreciated feature of online psychiatric research using self-report symptom surveys. In such surveys, participants rate their endorsement of various psychiatric symptoms and, since most individuals in the general population tend to endorse no or few symptoms in many symptom domains, the resulting ground-truth symptom distributions tend to be heavily positively skewed \cite{lowe2008validation, tomitaka2018distributional}. In this situation, the assumption that C/IE responding merely increases unsystematic measurement noise becomes untenable. Because of the positive skew in the ground-truth symptom distribution, participants who respond carelessly to the symptom survey are more likely to report higher levels of symptom endorsement relative to participants who complete the survey accurately \cite{chandler2020participant, ophir2020turker, king2018random}. Consequently, unless C/IE survey responses are carefully identified and removed, a considerable proportion of putatively symptomatic individuals in an online sample may, in fact, be participants who have not engaged with the experiment with sufficient attention or effort.

@@

Here, we demonstrate the real risk that C/IE responding can lead to spurious symptom-task correlations in computational psychiatry research. First, we asked to what extent recent studies in computational psychiatry screen participants based on self-report symptom data. We found that the majority of these studies did not screen participants' survey data at all, and that very few followed best-practice recommendations for survey data screening. We then asked whether behavioral screening alone was sufficient to identify participants engaging in C/IE responding on psychiatric symptom surveys. In two new datasets from two separate online labour marketplaces, we found that screening based on task behavior fails to completely identify participants engaging in C/IE responding on surveys. Lastly, we investigated whether, under these circumstances, C/IE responding led to spurious correlations between symptom severity and task performance for positively-skewed symptom measures. Consistent with the logic set out above, we confirmed that failure to appropriately screen out C/IE survey responding in the proof-of-concept datasets that we collected would have produced a number of spurious correlations between task behavior and self-reported symptoms that are abolished when data are screened more thoroughly.

@@

\begin{table}[!t]
\centering
\small
\setlength{\tabcolsep}{6pt}
\begin{tabular}{lcccccccc}
\toprule
& & \multicolumn{3}{c}{Total Score} & \multicolumn{2}{c}{Cronbach's $\alpha$} & \multicolumn{2}{c}{\% Clinical Cutoff} \\
\cmidrule(lr){3-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
 Subscale & Skew & Attentive & C/IE & \textit{t}-value & Attentive & C/IE & Before & After \\
\midrule
7-up   &  0.81 &    3.87 &   10.15 &  $-13.31^*$ &    0.84 &    0.84 &     13.0\% &       4.0\% \\
7-down &  0.76 &    4.75 &   10.68 &   $-9.99^*$ &    0.94 &    0.88 &     17.4\% &       9.3\% \\
GAD-7  &  0.75 &    4.86 &    9.73 &   $-7.88^*$ &    0.92 &    0.87 &     25.9\% &      17.3\% \\
BIS    &  0.78 &    7.74 &    7.92 &   -0.54 &    0.83 &    0.62 &        - &             - \\
BAS    &  0.17 &   15.73 &   16.16 &   -0.91 &    0.84 &    0.71 &        - &             - \\
SHAPS  &  0.26 &    8.02 &   10.85 &   $-4.04^*$ &    0.90 &    0.81 &     17.9\% &      14.6\% \\
PSWQ   &  0.19 &    4.78 &    6.74 &   $-4.78^*$ &    0.93 &    0.81 &      7.3\% &       7.0\% \\
\bottomrule
\end{tabular}
\captionsetup{width=0.88\textwidth}
\caption{Descriptive statistics of the self-report symptom measures between attentive and C/IE participants. Skew: the empirical skewness of the distribution of total symptom scores. Total score: the average symptom score across attentive and C/IE participants. Stars indicate statistical significance at $p<0.05$. Cronbach's $\alpha$: a measure of response consistency, where values closer to 1 indicate greater consistency in responses. \% Clinical Cutoff: the percentage of participants reaching threshold for clinical symptomology before and after screening based on the infrequency measure. (The BIS/BAS scales do not have clinical thresholds.)}
\label{tab:validation}
\end{table}

@@

\subsection{Pattern of results generalizes to alternative tasks, self-report measures, and quality assurance protocols}

One possible concern with the results presented so far is that they are specific to one specific instantiation of our experimental design. For example, with more stringent quality assurance protocols during participant recruitment, or perhaps a different task or set of self-report measures, one might wonder if spurious correlations would remain such a threat.

To evaluate the generalizability of our findings, we therefore conducted a conceptual replication experiment in which an independent sample of N=393 participants (N=193 from MTurk, N=200 from Prolific) completed a more difficult cognitive task, the two-step task \cite{daw2011model}, and an alternate set of self-report measures (see Appendix B for complete details). Importantly, participants were recruited \textit{after} CloudResearch and Prolific implemented new protocols to improve data quality on their respective platforms. Moreover, participants completed self-report symptom measures as before, but also personality measures with no hypothesized relationship to model-based planning behavior on the two-step task.

For the sake of brevity, only the main pattern of finding are reported here (the complete results are reported in Appendix B). In the replication sample, 55 out of 393 participants (14\%) endorsed a logically invalid or improbable response on one or more of the infrequency items when completing the self-report measures. This is roughly two-thirds of the number of participants who were flagged for C/IE responding in the original study, suggesting that the newer quality assurance protocols are at least partially effective. In the self-report symptom measures, we replicated the finding that total scores were noticeably exaggerated in participants suspected of C/IE responding only for the symptom measures where overall rates of symptom endorsement were the lowest. Similarly, we again found that task and self-report screening measures showed low correspondence; that is, excluding participants on the basis of poor behavioral performance was ineffective for identifying and removing participants who engaged in C/IE responding on self-report surveys. Finally, with no exclusions applied, we observed spurious correlations between performance on the two-step task and total scores, only for the self-report measures (both symptom and personality measures) with a mean-shift in scores between attentive participants and participants suspected of C/IE responding. In contrast with our original findings, however, we found that excluding participants based on self-report \textit{or} task screening measures was sufficient to ablate these spurious correlations.

In sum, we replicated most of the main findings from the original study in an independent sample of participants completing a novel set of tasks and self-report measures. Although we found that screening on task behavior was sufficient to protect against spurious correlations in the replication sample, it is difficult to generalize and predict when or why this might be the case for other datasets. As such, screening for C/IE responding in both task and self-report measures still remains the best approach to protect oneself against the possibility of spurious correlations.

\subsection{Individuals with and without diagnosed psychiatric disorders fail attention checks at similar rates}

One major concern with performing rigorous screening and exclusion of participants based on C/IE detection methods is that we might inadvertently introduce an overcontrol bias \cite{elwert2014endogenous}. That is, to this point we have treated the tendency towards C/IE responding as independent from psychopathology. However, to the extent that C/IE responding reflects lack of motivation \cite{barch2015mechanisms}, avoidance of effort \cite{cohen2001impairments, culbreth2016negative}, or more frequent lapses of attention \cite{kane2016individual, robison2017neurotic}, one might hypothesise a true underlying association between psychopathology and careless responding in online studies. In such cases, it is plausible that rigorous screening of C/IE responding might lead to the differential exclusion of truly symptomatic participants.

To explore this possibility, we embedded attention checks into the self-report measures of two studies of patients with major depressive disorder (see Appendix C for complete details). Specifically, a total of N=45 psychiatric patients and N=20 healthy controls completed a series of self-report symptom measures, online, on their computers from the comfort of their homes. (The psychiatric patients were confirmed to meet criteria for a diagnosis of major depressive disorder through a structured clinical interview.) In total, 16 of 65 (24.6\%) participants failed one or more attention checks. Subdivided by group, 6 of 20 (30\%) healthy participants and 10 of 45 (22\%) psychiatric participants were flagged for C/IE responding.

Using these data, we computed pairwise Bayes factors comparing three candidate models: attention check failure rates are equal between healthy and psychiatric participants (M1); failure rates are greater in psychiatric patients (M2); and failure rates are greater in healthy participants (M3). A model assuming equal rates of failure between healthy and psychiatric participants was 2.88 times more likely than the model assuming greater rates for psychiatric patients. In turn, a model assuming lower rates of failure for psychiatric patients was 1.27 times more likely than the model assuming equal rates. Finally, a model assuming lower rates of failure for psychiatric patients was 3.65 times more likely than the model assuming higher rates for psychiatric patients. Only the final comparison exceeds the cutoff value of 3, which is conventionally treated as the minimal amount of evidence required to treat a difference in model fit as meaningful. Though the small sample precludes any definitive conclusion, it is noteworthy that the model least consistent with the data is the one where psychiatric patients are more likely to fail infrequency item attention checks. These data suggest that it is unlikely that participants flagged for C/IE responding in the main analyses were disproportionately individuals with high psychiatric symptom severity. Accordingly, we conclude that our results are unlikely to be overly influenced by an overcontrol bias.

@@

One complicating factor for our argument is that C/IE responding may manifest in other ways than simply random responding for both behavioral tasks and self-report surveys. Indeed, there are more ways to respond carelessly than to respond attentively to a task or self-report inventory (e.g., random response selection, straight-lining, zig-zagging, acquiescence bias) \cite{curran2016methods}. The specific response strategy a participant adopts is likely to reflect the idiosyncratic integration of multiple perceived benefits (e.g. time saved, effort avoided) and costs (e.g. loss of performance bonuses, risk of detection and pay forfeited). As has been previously documented \cite{desimone2018dirty}, the presence of multiple response strategies makes it clear why certain screening measures are more or less likely to correlate. For example, the inter-item standard deviation and personal reliability measures are both sensitive to statistically random responding, but less sensitive to straight-lining. Most importantly, a diversity of heuristic response strategies highlights the need for many screening measures of C/IE responding, each sensitive to different heuristic strategies.

@@

One objection to the rigorous screening and exclusion of participants based on C/IE detection methods is that we might inadvertently introduce an overcontrol bias. That is, to the extent that C/IE responding might reflect symptoms common to psychopathology (e.g. low motivation, effort avoidance, inattentiveness), rigorous screening of C/IE responding might lead to the differential exclusion of truly symptomatic participants. To explore this possibility, we embedded attention checks into the self-report measures of two studies of patients with major depressive disorder. Though our final sample was small, we did not find evidence for the proposition that patients were more likely to fail attention checks than healthy controls (if anything, non-psychiatric participants were more likely to be flagged by C/IE screening). These results provide preliminary evidence that rigorous C/IE screening is unlikely to result in overcontrol bias.

Given that the results of our patient study are preliminary and warrant further investigation, researchers might still be wary of possible overcontrol bias. However, when using self-report questionnaires for screening, for overcontrol to seriously impact results it would have to be the case that symptomatic participants frequently endorse improbable or impossible responses to infrequency-item checks (e.g., responding `Agree' to ``I competed in the 1917 Olympic Games"). In this case, and even if such participants truly are experiencing severe symptoms of motivation or attention, there is likely to be limited utility in measuring these symptoms using a self-report measure that they are unable to complete veridically. A similar rationale underlies the widespread use of semi-structured interviews and other clinician-report measures rather than self-report measures for in-clinic psychiatric research. We would therefore argue that, if the psychiatric phenomenon being studied is such that this issue warrants concern, the research question may be better suited to an in-person study design involving participants in the clinic who meet full diagnostic criteria than a correlational design involving an online convenience sample.

@@

409 total participants were recruited to participate in an online behavioral experiment in late June - early July, 2020. Specifically, 208 participants were recruited from Amazon Mechanical Turk (MTurk) and 201 participants were recruited from Prolific. This study was approved by the Institutional Review Board of Princeton University (\#5291), and all participants provided informed consent. Total study duration was approximately 10 minutes per participant. Participants received monetary compensation for their time (rate USD \$12/hr), plus an incentive-compatible bonus up to \$0.25 based on task performance.

@@

@@

\section*{References}

\printbibliography[heading=main]
\end{refsection}

\break

@@

\begin{refsection}[supp]
\section*{Appendix A}
\setcounter{figure}{0}
\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\thefigure}{S\arabic{figure}}

\subsection*{Posterior predictive checks}

@@

\subsection*{C/IE responding manifests as qualitatively distinct behavioral strategy}

In an exploratory analysis, we employed a theory-agnostic modeling approach to investigate how C/IE participants on the probabilistic reversal-learning task compared to attentive participants. The motivation for this analysis was to better understand why C/IE responding was inconsistently predicted by chance-level performance, and also correlated with asymmetric learning rates.

To characterize participants' choice behavior, we adapted the softmax regression model from \cite{seymour2012serotonin}. This model estimates, for each participant, how much their choice depends on the recent history of trial events (rewarding outcomes, non-rewarding outcomes, and choices from the preceding 5 trials). Specifically, the influence of the history of particular type of event is defined as:

where $x_{t-i}$ is a binary indicator [0,1] denoting if an event (i.e. reward, non-reward, previous choice) occurred on trial $t-i$ and $w_{t-1}$ is the associated decision weight. These weights were estimated for rewards, non-rewards, and previous choices up to five trials in the past. The overall tendency to choose a particular choice option is dictated by a softmax choice rule:

We fit the softmax regression model using Stan following the same procedure as for the theory-based analyses. Participant parameters were fit individually (i.e. not hierarchically)  so as to prevent bias during parameter estimation from partial-pooling between attentive and C/IE participants. Parameters were sampled with Gaussian priors with $\mu = 0$ and $\sigma = 5$.

The regression weights for each event, averaged within attentive and C/IE participants, are presented in Figure \ref{fig:regression}. Comparing attentive to C/IE participants, we observed a credible difference (i.e. 95\% highest density intervals excluded zero) only for the $T-1$ weight for previous choice. That is, attentive participants were more likely to repeat their previous choice (i.e. greater choice hysteresis) than were C/IE participants.

\begin{figure}[!t]
\includegraphics[width=16cm]{../figures/main_03c.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Softmax regression decision weights in attentive (red) and C/IE (blue) participants. The weights dictate the extent to which the recent history of rewards, nonrewards, or previous choices influence current choice. Error bars indicate the 95\% highest density interval of each weight. * denotes where the 95\% highest density interval of the difference in weights excluded zero.}
\label{fig:regression}
\end{figure}

\break
\section*{Appendix B: Replication study}

\subsection*{Background \& motivation}

Here we report a conceptual replication of our original study. The motivations for conducting a replication study were threefold. First, we wanted to examine the generalizability of our findings under new labor platform conditions. Since summer 2020, when the original data were collected, online labor markets like Mechanical Turk/Cloudresearch and Prolific Academic have undergone important changes. To address rampant data quality issues on Mechanical Turk, Cloudresearch introduced their ``Approved Participants'' filter. When selected, only Mechanical Turk participants with a prior history of attentive and careful work are invited to participate in experiments \cite{cloudresearch_2020-2, hauser2021evaluating}. Similarly, a deluge of new, lower-quality users signed up to participate in studies on Prolific in summer, 2021 \cite{Letzter2021-bw}. In response, Prolific introduced new controls and filters to improve data quality on the platform \cite{prolific_2021}. In the wake of these changes, we wanted to explore the relevance of our original findings. Specifically, with these new safeguards, we wanted to examine whether the chance of spurious correlations has been seriously lessened.

Second, we wanted to examine the generalizability of our findings to other behavioral measures. In the original study, we used a short, straightforward, and relatively easy reversal-learning task. Our motivation then was to demonstrate the risk of spurious correlations even for experiments where the possibility of participant fatigue (and consequently C/IE responding) had been minimized. One consequence of this design, however, was that the majority of participants performed reasonably well on the task (i.e. only 26 participants, or 7\% of the sample, exhibited choice accuracy at or below statistically chance levels). This fact may in part explain why we observed such low correspondence between self-report and task-based screening measures. As such, we wanted to repeat our experiment and analyses using a more difficult task. Therefore, in the replication study we used the two-step task \cite{daw2011model}, which is both more challenging and lengthy than our original reversal-learning task.

Finally, we wanted to examine the generalizability of our findings to other self-report measures. In the original study, we found that, all else being equal, self-report symptom measures with low rates of endorsement were more likely to yield spurious correlations with behavior in the presence of C/IE responding. This effect should not be limited to symptom measures, however, and should extend to any self-report measure with an expected skewed or asymmetric score distribution. Therefore, in the current replication study, we use two sets of self-report scales: one set of psychiatric symptom measures and one set of personality measures. As before, both sets include scales whose score distributions are expected to be symmetric or asymmetric (skewed). Crucially, the two personality scales measure constructs (i.e. artistic interests, greed avoidance) that should have no meaningful relationship with model-based choice behavior on the two-step task. Therefore, the personality measures serve as a stronger test of our hypothesis that spurious correlations between self-report and behavioral measures are more likely for skewed score distributions.

\subsection*{Methods}

\subsubsection*{Participants}

400 total participants were recruited to participate in an online behavioral experiment in February, 2022. Specifically, 200 participants were recruited from Amazon Mechanical Turk (MTurk) and 200 participants were recruited from Prolific. This study was approved by the Institutional Review Board of Princeton University (\#11968), and all participants provided informed consent. Total study duration was approximately 20 minutes per participant. Participants received monetary compensation for their time (rate USD \$12/hr), plus an incentive-compatible bonus up to \$1.00 based on task performance.

Participants were eligible if they resided in the United States or Canada. Participants from MTurk were recruited with the aid of CloudResearch services \cite{litman2017turkprime-2} using their ``Approved participants''  data quality filters \cite{cloudresearch_2020-2}. As in the original study, MTurk workers were not excluded based on work approval rate or number of previous jobs approved \cite{robinson2019tapped-2}. No other exclusion criteria were applied during recruitment.

The data from N=7 participants who completed the experiment were excluded prior to analysis. These participants (all from MTurk) disclosed that they had also completed the experiment on the other platform were excluded. This left a final sample of N=393 participants (MTurk: N=193, Prolific: N=200) for analysis. The demographics of the sample split by labour market is provided in Table \ref{tab:demographics_02}. The participants recruited from MTurk were older on average ($M = 4.9 \ \text{yrs}, t = 4.248, p < 0.001$) and comprised of fewer women ($z = 4.248, p < 0.001$).

\begin{table}[!h]
\centering
\small
\setlength{\tabcolsep}{15pt}
\begin{tabular}{ lrrrr }
\toprule
& \multicolumn{2}{c}{MTurk} & \multicolumn{2}{c}{Prolific} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5}
Total & \multicolumn{2}{c}{N=193} & \multicolumn{2}{c}{N=200} \\
\midrule
Age & N & \% & N & \% \\
\cmidrule(lr){1-5}
18-25   &    11 &   5.7 &     47 &  23.5 \\
26-35   &    71 &  36.8 &     76 &  38.0 \\
36-45   &    60 &  31.1 &     41 &  20.5 \\
46-55   &    29 &  15.0 &     22 &  11.0 \\
55+     &    22 &  11.4 &     14 &   7.0 \\
\midrule
Gender & N & \% & N & \% \\
\cmidrule(lr){1-5}
Female &      68 &  35.2 &      122 &  61.0 \\
Male   &     124 &  64.2 &       73 &  36.5 \\
Other  &       1 &   0.5 &        5 &   2.5 \\
\midrule
Ethnicity & N & \% & N & \% \\
\cmidrule(lr){1-5}
Hispanic or Latino     &    16 &   8.3 &     12 &   6.0 \\
Not Hispanic or Latino &   177 &  91.7 &    180 &  90.0 \\
Rather not say         &     0 &   0.0 &      8 &   4.0 \\
\midrule
Race & N & \% & N & \% \\
\cmidrule(lr){1-5}
American Indian/Alaska Native &     2 &   1.0 &      0 &   0.0 \\
Asian                         &    15 &   7.8 &     41 &  20.5 \\
Black or African American     &    13 &   6.7 &     12 &   6.0 \\
White                         &   156 &  80.8 &    133 &  66.5 \\
Multiracial                   &     6 &   3.1 &     10 &   5.0 \\
Rather not say                &     1 &   0.5 &      4 &   2.0 \\
\bottomrule
\end{tabular}
\captionsetup{width=0.88\textwidth}
\caption{The demographics of each sample by online labour market.}
\label{tab:demographics_02}
\end{table}

\subsubsection*{Experiment}

Participants completed a gamified version of the two-step task \cite{daw2011model} designed to dissociate ``model-free'' and ``model-based'' decision-making. On every trial of the task, participants were presented with two choice options that led probabilistically to either of two second-stage ``states''. From there, participants made an additional choice between two options that were associated with a unique probability of returning either a reward (1 point) or a non-reward (0 points). Crucially, each first-stage choice option led mostly (70\% of the time) to one of the two second-stage states; this probability was fixed throughout the experiment. The probabilities of reward for each of the four second-stage options changed slowly and independently, according to Gaussian random walks. Participants completed 201 trials of the task, with an optional break after the first half.

As a cover story, the two-step task was introduced to participants as a ``space treasure'' game \cite{decker2016creatures}. In the game, participants made choices between two different colored rocket ships (first stage choice) that probabilistically took them to one of two planets. On each planet, there were two alien creatures to ``trade'' with. If chosen as a trading partner, an alien would either give ``treasure'' (1 point) or ``junk'' (0 points). Participants were told their goal was to collect as much treasure as possible. Participants were informed that the rocket ships mostly traveled to one planet, but sometimes went to the other. Participants were also instructed that the chance an alien would give them treasure would change slowly over the course of the task. Prior to beginning the experiment, participants had to correctly answer three sets of comprehension questions about the instructions. Failing to correctly answer all items forced the participant to start a section of the instructions over.

The task was programmed in jsPsych \cite{deleeuw_2015_jspsych-2} and distributed using custom web-application software. The experiment code is available at \url{https://github.com/nivlab/sciops}, and the web-software is available at \url{https://github.com/nivlab/nivturk}. A playable demo of the task is available at \url{https://nivlab.github.io/jspsych-demos/tasks/two-step/experiment.html}.

\subsubsection*{Self-report measures}

Prior to the start of the two-step task, participants completed four self-report measures (in a randomized order). The first was the 14-item seven-up/seven-down scale (7u/7d; \cite{youngstrom_2013_susd-2}), which measures lifetime incidence of depressive and (hypo)mania symptoms. These scales are expected to elicit lower rates of symptom endorsement, thereby resulting in asymmetric (right-skewed) score distributions. Participants also completed a 7-item measure of general anxiety symptoms over the last year (e.g. ``I was overwhelmed by anxiety.''; \cite{watson2022development}). This scale is expected to elicit moderate rates of symptom endorsement, thereby resulting in symmetric score distributions. We expected the depression and mania measures to be at greater risk than the anxiety measure for spurious correlations with behavior on the two-step task.

In addition, participants completed a 6-item measure of artistic interests (e.g. ``I believe in the importance of art''; \cite{ashton2007empirical}). Based on previous studies, this scale is expected to elicit high rates of endorsement, thereby resulting in an asymmetric (left-skewed) score distribution. Finally, participants completed a 6-item measure of greed avoidance, which measures tendencies towards status-seeking (e.g. ``I am out for my own personal gain''; \cite{ashton2007empirical}). Based on previous studies, this scale is expected to elicit moderate rates of endorsement, thereby resulting in a symmetric score distribution. We  expected the artistic interests scale to be at greater risk than the greed avoidance scale for spurious correlations with behavior on the two-step task.

\subsubsection*{Correspondence of screening measures}

As in the original study, we measured the correspondence of screening measures based on the task and self-report behavior. We calculated a number of standard measures of data quality from each participant's task behavior (four in total) and self-report responses (five in total). The self-report screening measures were identical to those used in the original study, except that we used (mostly) new infrequency items. We describe each of the new screening measures below.

\begin{enumerate}
  \item Have there been times in your life where you blinked your eyes at least once per day?
  \item Have there been times of a couple days or more when you were able to breathe underwater (without an oxygen tank)?
  \item I was worried about the canine World Cup.
  \item I have used a computer.
\end{enumerate}

Prior to conducting the study, the infrequency items were piloted on an independent sample of participants to ensure that they elicited one dominant response. We also included one instructed item (``Please select "Neutral" as your response'') to compare to the infrequency items. We measured the number of suspicious responses made by each participant to these questions. For thresholded analyses, participants were flagged if they responded incorrectly to one or more of these items.

\subparagraph{Key variability} Key variability was defined as the fraction of trials a participant pressed the left arrow key across first stage choices during the two-step task. Key variability could range from 0.00 (only right arrow key used) to 1.00 (only left arrow key used). Extreme values (i.e. closer to zero or one) are indicative of more careless responding during the task.

\subparagraph{Choice variability} Choice variability was defined as the fraction of trials a participant chose the same first-stage choice option during the two-step task. (The left-right position of first-stage choice options were randomized across trials, such that response key and response choice were orthogonalized.) Choice variability could range from 0.00 (only selected the first option) to 1.00 (only selected the second option). Extreme values (i.e. closer to zero or one) are indicative of more careless responding during the task.
